---
title: "W203 Lab01 Broadband Exploration"
output: pdf_document
author: Chandra Sekar, Bhuvnesh Sharma, Eugene Tang
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
#setwd("C:/W203/broadband") # change as necessary
setwd("~/Desktop/EugeneTang/Grad School/Berkeley/W203_Statistics/Homework/Lab01/broadband/") # change as necessary
```

## Introduction

### Research Question
With the recent rise of the Internet and its increasing importance in our lives, the availablility of access to the Internet has also become an increasingly large question and necessity in some societies. In this analysis, we look at broadband markets and in particular, three aspects of these markets and their relationships with each other: 

* Price: how much does it cost to access the Internet
* Penetration: what fraction of customers have access to network service
* Speed: what rate can customers upload or download bits of data

We in particular consider this data in the context of open access policies. Much of the developed world has developed aggressive regulatory structures to compel network owners to increase penetration while there are some nations that do not. It is still an open debate on whether such policies are beneficial or harmful in price, penetration, and speed.

In this analysis we seek to tackle two main questions: 

* Does a trade-off exist between network price, penetration, and speed?
* Is there evidence for beneifical effects of open access policies?

### Dataset Setup (code)
Here is the code we used to prepare our dataset. Please see the below sections to see why certain decisions were made in the preparation of the dataset.

```{r}
library(dplyr)
library(car)
library(scatterplot3d)

df_price = read.table("Price.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
df_penetration = head(read.table("Penetration.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE), -1) # last row is empty
df_speed = head(read.table("Speed.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE), -2) # last two rows is empty

# some dataset cleaning (see "Data Processing" for more details on why we did this)
drops_penetration <- c("X") # extra column to remove
df_penetration = df_penetration[ , !(names(df_penetration) %in% drops_penetration)]
colnames(df_speed)[2] <- "Country.Code" # rename country code column

# convert numeric fields to the numeric type
convert_to_numeric = function(col) {
  return(as.numeric(sub("[\\$%,]","", col)))
}
NON_ID_DATA_START <- 3 # columns 1 and 2 are the country and country code data
df_price[NON_ID_DATA_START:length(df_price)] = lapply(df_price[NON_ID_DATA_START:length(df_price)], convert_to_numeric)
df_penetration[NON_ID_DATA_START:length(df_penetration)] = 
  lapply(df_penetration[NON_ID_DATA_START:length(df_penetration)], convert_to_numeric)
df_speed[NON_ID_DATA_START:length(df_speed)] = 
  lapply(df_speed[NON_ID_DATA_START:length(df_speed)], convert_to_numeric)

# Join the data
df_partial = full_join(df_penetration, df_price, by = c("Country", "Country.Code"))
df_full = full_join(df_partial, df_speed, by = c("Country", "Country.Code"))
```

### Dataset Description
Our dataset comes in three csv files. One for price, penetration, and speed respectively. Each dataset contains observations on 30 countries, with one row for each country. Each dataset contains a variety of variables. Below we include tables of each variable, its type, and its description.

Each of the three datasets contains string columns to represent country and country code field. We use the two fields to join the datasets together since they are unique across each row (though either field individually would also have sufficed. For conciseness, we exclude these two columns in the tables below.

```{r}
writeLines(paste(length(df_full$Country), "countries"))
writeLines(paste(df_full$Country, collapse = ", ", sep = ""))
```

#### Price Dataset
Our price dataset contained information on the cost to access different levels of Internet

Column                               | Interpretation
------------------------------------ | ------------------------------------
Price for low speeds, combined       | Price (in USD$) for low-speed Internet
Price for med speeds, combined       | Price (in USD$) for medium-speed Internet
Price for high speeds, combined      | Price (in USD$) for high-speed Internet
Price for very high speeds, combined | Price (in USD$) for very high-speed Internet

#### Penetration Dataset
Our penetration dataset contained information on how widespread network availability is in a country.

Column                               | Interpretation
------------------------------------ | ------------------------------------
Penetration per 100 OECD, 2008       | "Penetration" per 100 people in 2008 as measured by the OECD
Penetration per 100 OECD, 2007       | "Penetration" per 100 people in 2007 as measured by the OECD
Household penetration, OECD          | Household penetration as measured by the OECD
2G and 3G penetration per 100, OECD  | 2G and 3G network penetration per 100 people for OECD
Penetration per 100 GC               | Penetration per 100 people as measured by GC
3G penetration per 100               | 3G network penetration per 100 people
Growth in 3G penetration             | Growth in 3G network penetration
Wi-Fi hotspots, JiWire               | Number of Wi-Fi hotspots
Wi-Fi hotspots per 100,000, JiWire   | Number of Wi-Fi hotspots per 100,000 people
Percent of population in urban areas | Percent of population in urban areas with access to the Internet

#### Speed Dataset
The speed dataset has various metrics on the speed of the network available in a country.

Column                               | Interpretation
------------------------------------ | ------------------------------------
Maximum advertised speed OECD (kbps)       | (same as column name)
Average advertised speed OECD (kbps)       | (same as column name)
Average actual speed, Akamai (kbps)        | (same as column name)
Average download speedtest.net (kbps)      | (same as column name)
Standard deviation download, speedtest.net | (same as column name)
Average upload speedtest.net (kbps)        | (same as column name)
Standard deviation upload, speedtest.net   | (same as column name)
Average latency speedtest.net              | Average latency in ms (measured by speedtest.net)
Standard deviation latency, speedtest.net  | Standard deviation of latency in ms (measured by speedtest.net)
Median download, speedtest.net (kbps)      | (same as column name)
Median upload, speedtest.net (kbps)        | (same as column name)
Median latency, speedtest.net              | Median latency in ms (measured by speedtest.net)
90p. Download, speedtest.net (kbps)        | Top 90% of download speed (measured by speedtest.net)
90p. Upload, speedtest.net (kbps)          | Top 90% of upload speed (measured by speedtest.net)
10p. Latency, speedtest.net                | Lowest 10% of latency in ms (measured by speedtest.net)

### Data Processing / Preparation
When first starting with the dataset, we noticed that two of the files (Speed.csv and Penetration.csv) had extra lines at the bottom. To help work with this, we manually removed the two lines and saved the results in Speed_Modified.csv and Penetration_Modified.csv (also attached). 

The next thing we noticed was that the penetration dataset had an extra column called X with no values. We chose to removed this column since it has no values. In the speed dataset, we noticed that the country code column was named differently ("Country code") from that in the penetration and price datasets ("Country.Code"). To adjust for this, we renamed the country code column to "Country.Code" to help facilitate joins later on.

```{r eval=FALSE}
# some dataset cleaning
drops_penetration <- c("X") # extra column
df_penetration = df_penetration[ , !(names(df_penetration) %in% drops_penetration)]
colnames(df_speed)[2] <- "Country.Code"
```

After performing these adjustments, we found that a lot of data were read as factors because some columns had percentage signs or commas in the numbers which had the read.tables function interpret the input as Strings instead of numbers. Because we would want to interact with all of the values as numbers (short of the country and country code columns), we next converted all of the columns to numeric type.

```{r eval=FALSE}
# convert numeric fields to the numeric type
convert_to_numeric = function(col) {
  return(as.numeric(sub("[\\$%,]","", col)))
}
NON_ID_DATA_START <- 3 # columns 1 and 2 are the country and country code data
df_price[NON_ID_DATA_START:length(df_price)] =
  lapply(df_price[NON_ID_DATA_START:length(df_price)], convert_to_numeric)
df_penetration[NON_ID_DATA_START:length(df_penetration)] =
  lapply(df_penetration[NON_ID_DATA_START:length(df_penetration)], convert_to_numeric)
df_speed[NON_ID_DATA_START:length(df_speed)] = lapply(df_speed[NON_ID_DATA_START:length(df_speed)], convert_to_numeric)
```

Finally, we joined the three tables to make it easier to work with.
```{r eval=FALSE}
# Join the data
df_partial = full_join(df_penetration, df_price, by = c("Country", "Country.Code"))
df_full = full_join(df_partial, df_speed, by = c("Country", "Country.Code"))
```

### Data Quality Evaluation
There were a few columns that had fewer than 30 observations.

```{r}
cols_with_nas <- colnames(df_full)[colSums(is.na(df_full)) > 0]
```

Of these columns, the measurement of the price of very high Internet speeds by far was the most incomplete (had the most NAs). One possible explanation for this is that not all countries have very high speed Internet and thus would not have a value for this column. For columns with NAs, we decided to omit the countries with the NAs from our analyses since we do not have a good default to put in.

```{r}
print_nas = function(colname) {
  writeLines(paste("Number of NAs in ", colname, ": ", sum(is.na(df_full[colname])), sep = ""))
}
invisible(lapply(cols_with_nas, print_nas))
```

Outside of this, our data was fairly complete. However, looking through our data, we did notice some irregularities. The first we found is that Poland's value for the percent of population in urban areas field is 162%. We do not have a full understanding of what that means, but it seems that this field measures a percentage of the population of people in urban areas, so since 162% > 100%, we are a little dubious of that data point and feel that it could be likely due to reporting error. For this reason, we decided to remove Poland from analyses containing this column.

Another irregularity was that Turkey's 3G penetration per 100 is 0 despite having 20.20% growth in 3G penetration. It is possible that Turkey had a really really low 3G penetration such that a 20.20% growth would still result in an average of 0 per 100. However, given some of the other values in the other fields, (e.g. 2G and 3G penetration per 100, OECD having a value of 84.93), we are a little dubious of this value and decided to remove Turkey from analyses containing this column. 

Finally, we noticed that in the measurement of the standard deviation of latency for Iceland, Iceland has an unusually high value of 1,199 (where all the other standard deviations are around 300). Further irregularities in the data we found and how we handle them can be found in the following sections.

## Univariate Analysis of Key Variables
Now we have the combined data in the data frame (df_full), we can use the invividual dataframe to perform the univariate analysis of key variables. We renamed the column names for clarity while displaying summary and the plots.

### Price Dataset
```{r}
colnames(df_price)[NON_ID_DATA_START:length(df_price)] <- c("Low_Speed", "Medium_Speed", "High_Speed", "VeryHigh_Speed")
summary(df_price)
```

By looking at the min, max and the different quartiles we can see there are some outliers. Using a boxplot we can clearly see which variable has outliers and a sense for how much deviation there is.

```{r fig.width = 8}
boxplot(df_price[NON_ID_DATA_START:length(df_price)],ylab = "Price in USD" , col = c("palevioletred","sienna","royalblue","red1"), main="Boxplots of prices for different levels of Internet")
```

The max value in the "High Speed" variable is \$210.36 for the country Poland, whereas the mean for the "High Speed" variable is \$55 and the 3rd Quartile is \$64. This seems a bit irregular, though it could be possible. The min value for the "High speed" vaiable is \$0.69 for the country Luxembourg which could be wrong since that country's Low speed and Medium speed prices are 26 and 36 dollars respectively.

Looking at the NAs, there are 11 NAs for the "VeryHigh Speed" which could mean that those 11 countries do not have veryhigh speed broadband services. But the NA in "Low Speed" variable for the country Belgium is something we need to be aware of when doing further analysis. Similarly, Mexico and Turkey have NAs in the "High Speed" variable.

Since the "Low Speed" and "VeryHigh Speed" variables are incomplete, and given the likely chance that the "High Speed" variable has bad data, we decided to use the price of the "Medium Speed" Internet as a proxy for price in further analysis. In addition, we note that the "Medium Speed" variable is fairly highly correlated with the other three variables:

```{r}
cor(df_price$Low_Speed, df_price$Medium_Speed, use="pairwise.complete.obs")
cor(df_price$High_Speed, df_price$Medium_Speed, use="pairwise.complete.obs")
cor(df_price$VeryHigh_Speed, df_price$Medium_Speed, use="pairwise.complete.obs")
```

### Speed Dataset
```{r}
colnames(df_speed)[NON_ID_DATA_START:length(df_speed)] <- c("Max Advertised", "Average Advertised", "Average Actual", "Average Download", "SD download", "Average upload", "SD upload", "Average Latency", "SD Latency", "Median Download", "Median upload", "Median Latency", "Top 90% Download", "Top 90% Upload", "Bottom 10% Latency")
summary(df_speed)
```

One of the key variable in this dataset is the "Average Actual speed". This has NAs for the countries Hungary, Poland, Slovak and Turkey. This is interesting because these countries do have values in the average download speed, upload speed and other variables. Also to note are the Max value of this variable, 15,239Kbps for Korea and the min value 948Kbps for Mexico. 

```{r}
hist(df_speed$`Average Actual`, breaks=20, col="red", main = 'Histogram of Average Actual Speed, Akamai', xlab = 'Average Actual Speed, Akamai (kbps)')
```

The other Key variable in this dataset is "Median Download speed"

```{r}
hist(df_speed$`Median Download`, breaks=20, col="royalblue1", main = 'Histogram of Median Download Speed', xlab = 'Median Download Speed, speedtest.net (kbps)')
```

While we would like to use Average Actual Speed as a proxy for speed, it has some missing values. Looking at the above histogram and its similarity to the histogram of the "Average Actual Speed", we have decided to use "Median Download Speed" as the proxy speed variable for further analysis. As a further confirmation of this, we find that of Download Speed, Upload Speed, and Latency, Median Download Speed has a high (and the highest) correlation with Average Actual Speed. We also choose Median Download Speed rather than Average Download Speed because it is more robust to outliers.

```{r}
cor(df_speed$`Average Actual`, df_speed$`Median Download`, use="pairwise.complete.obs")
cor(df_speed$`Average Actual`, df_speed$`Median upload`, use="pairwise.complete.obs")
cor(df_speed$`Average Actual`, df_speed$`Median Latency`, use="pairwise.complete.obs")
```

While briefly looking at the other key variables we noticed that in the measurement of the standard deviation of latency for Iceland, Iceland has an unusually high value of 1,199 (where all the other standard deviations are around 300). The standard deviation for upload speed is 12,098 for Sweden. Again unsually high while the others have SD in the range of 4,000. Given these potential outliers, we need to keep them in mind when analyzing the rest of the data and how they can affect our analysis or skew our results (e.g. heavily influence the mean or line of best fit).

Below we also show the countries with the maximum median download and upload speeds, and minimum latency. We chose to use the median to avoid having outliers influence the results:

```{r}
df_speed$Country[which.max(df_speed$`Median Download`)]
df_speed$Country[which.max(df_speed$`Median upload`)]
df_speed$Country[which.max(df_speed$`Median Latency`)]
```

### Penetration dataset
```{r}
colnames(df_penetration)[NON_ID_DATA_START:length(df_penetration)] <- c("Penetration 2008", "Penetration 2007", "Household Penetration" ,"2G and 3G Penetration", "Penetration GC", "3G Penetration", "Growth in 3G" , "WiFi hotspots" , "WiFi per 100000", "Percent Urban" )
summary(df_penetration)
```

Poland's value for the percent of population in urban areas field is 162%. We do not have a full understanding of what that means, but it seems that this field measures a percentage of the population of people in urban areas, so since 162% > 100%, we are a little dubious of that data point and feel that it could be likely due to reporting error. For this reason, we decided to remove Poland from analyses containing this column.

Another irregularity was that Turkey's 3G penetration per 100 is 0 despite having 20.20% growth in 3G penetration.
Let's look at the boxplot to understand more abnormal values:

```{r}
boxplot(df_penetration[NON_ID_DATA_START:8], las=2, names=c("2008", "2007", "Household" , "2G & 3G", "GC", "3G"), main="Boxplots of key penetration variables")
```

```{r}
cor(df_penetration$`Penetration 2008`, df_penetration$`Penetration 2007`, use="pairwise.complete.obs")
cor(df_penetration$`Penetration 2008`, df_penetration$`Penetration GC`, use="pairwise.complete.obs")
cor(df_penetration$`Penetration 2007`, df_penetration$`Penetration GC`, use="pairwise.complete.obs")
```

The above boxplot and the correlation figures show that the penetration data for 2008, 2007 and GC are highly correlated and and can be used as is for further analysis. We chose the 2008 OECD penetration data as the proxy for the penetration in the key relationship analysis. 

Let's look at the boxplot of the remaining variables.

```{r}
boxplot(df_penetration[9], ylab = "Percentage" , xlab = "Growth in 3G", main="Boxplot of Growth in 3G by percentage")
```

The above plot of Growth in 3G has 2 values outside of the range.Australia and Austria have 510% and 339% of growth respectively.

```{r}
boxplot(df_penetration[10], ylab = "Total number" , xlab = "WiFi Hotspots", main="Boxplot of Total number of WiFi Hotspots")
```

Most of the countries have WiFi hotspots in the range of few hundreds to 5000, whereas Unitedstates has 67718 hotspots followed by France and UK in the 25 thousand range. Another interesting observation is Iceland has only 6. 

```{r}
boxplot(df_penetration[11], ylab = "Number per 100000 people" , xlab = "WiFi Hotspots", main="Boxplot of WiFi Hotspots per 100000 people")
```

In the above plot Sweden and Switzerland have the most hotspots for it's population, 74 and 70 respectively. Again on the low and possibly on the erroneous side is Iceland with 2 and Turkey with 0.6.

```{r}
boxplot(df_penetration[12], ylab = "Percentage" , xlab = "Urban Population", main="Boxplot of Urban population in percentage")
```

In the above plot the outlier or the questionable entry is for Poland with 162%. The rest of data is within the range of a typical unrban population percentage with a median of 77%

## Analysis of Key Relationships
###Price v. Speed v. Penetration
(TODO: remove poland?)
In order to perform the Key relationship analysis, we chose the "price for med speeds" from the Price dataset, "2008 Penetration OECD" from the Penetration dataset and "Median Download Speed" from the speed dataset for the reasons described in the previous section.

```{r}
scatterplotMatrix(~Price.for.med.speeds..combined + Penetration.per.100.OECD..2008 + Median.download..speedtest.net..kbps., data=df_full)
```

First Observation: When penetration increases the price comes down

```{r}
pairs(~Price.for.med.speeds..combined + Penetration.per.100.OECD..2008 + Median.download..speedtest.net..kbps., data=df_full)
```

Second Observation: Interesting observation is that download speed is higher at a lower price point, whereas the download speed doesn't increase when the price is higher.

Third Observation: As the penetration increases the download speed increases.

In order to analyze the key relationship between all 3 datasets, we used a 3dscatterplot. 

```{r}
scatterplot3d(df_full$Penetration.per.100.OECD..2008,df_full$Price.for.med.speeds..combined,log10(df_full$Average.download.speedtest.net..kbps.), pch=16, highlight.3d=TRUE,
  type="h", main="3D Scatterplot of Price Vs Speed Vs Penetration")
```

Threeway Relation: when the Penetration increases, it brings down the price and also increases the download speed.

###Median v. Average
One other interesting comparison to make is between the median and average speeds in the speed dataset. We would expect them to be highly correlated. We present the results below:

```{r}
cor(df_full$Average.download.speedtest.net..kbps., df_full$Median.download..speedtest.net..kbps., use="pairwise.complete.obs")
cor(df_full$Average.upload.speedtest.net..kbps., df_full$Median.upload..speedtest.net..kbps., use="pairwise.complete.obs")
cor(df_full$Average.latency.speedtest.net, df_full$Median.latency..speedtest.net, use="pairwise.complete.obs")
```

It is interesting that the correlation between average and median latency is 0.65. We create a scatterplot to look at the relationship between the two:

```{r}
plot(~Average.latency.speedtest.net + Median.latency..speedtest.net, data=df_full)
```

It looks like there is an outlier in the data. Poking through the data, we find this outlier is the country (TODO: XX). We do not use latency as a proxy above, but it is interesting to note how they are so different. Finally, here we also see that the average latency tends to be higher than the median, meaning that the data is left-skewed.

## Analysis of Secondary Effects
Whether a country has open access laws can also influence the appearance of penetration, price, speed, and their relationship. Assuming that we will be using the price to access Medium Speed Internet as a proxy for price, the median download speed as a proxy for speed, and the OECD 2008 penetration metric as a proxy for penetration, we take a quick glance at the difference between countries with and without open access laws.

We take the list of countries without open access laws from the provided assignment document:
```{r}
COUNTRIES_NO_OPEN_ACCESS = c("US", "MX", "SK")
df_full_wo_open_access = df_full[df_full$Country.Code %in% COUNTRIES_NO_OPEN_ACCESS, ]
df_full_with_open_access = df_full[!(df_full$Country.Code %in% COUNTRIES_NO_OPEN_ACCESS), ]

writeLines("Summary data for countries with open access")
summary(df_full_with_open_access[c("Price.for.med.speeds..combined", "Penetration.per.100.OECD..2008", "Median.download..speedtest.net..kbps.")])
writeLines("Summary data for countries without open access")
summary(df_full_wo_open_access[c("Price.for.med.speeds..combined", "Penetration.per.100.OECD..2008", "Median.download..speedtest.net..kbps.")])
```

First, we note that the countries without open access laws tend to have higher median price to access Internet, lower median penetration, and lower median download rates than countries with open access laws. This seems to correspond with theory, where countries with open access laws would be more motivated to increase penetration and decrease the price of the Internet so that more people can access it. At the same time, it might be that open access laws encouraged Internet providers to share infrastructure, thus leading to less waste and better Internet service as well (e.g. higher download speeds). However, we note that there are only three countries in our dataset without open access laws, so all of these results must be taken with a grain of salt and further analysis would be necessary before making any serious claims. 

Looking through this data, we also noticed that there could be many secondary effects other in play. The first is culture. For example, South Korea has a large culture of online gaming. This increases the demand for high speed Internet and could be a reason why its median download speed is the highest. Such secondary effects could confound the relationship between price, penetration, and speed, as well as whether open access laws make a difference. For example, the magnitude of price, penetration, and speed could be a function of South Korea's gaming-heavy culture rather than open access laws. Another possible secondary effect is the wealth of the country. Wealthier countries can afford more infrastructure and possibly even subsidize Internet costs. This would also be a driver for bringing down Internet prices, increasing penetration, and increasing Internet speed. Thus rather than open access laws influencing availability of Internet, it could be purely a function of wealth. Given these possible secondary variables, we would take these results with a grain of salt and look at more data before making any conclusions. I would try to control for certain variables such as a nation's wealth in future steps of this analyses.

## Conclusion
From our analysis of the 